The imprint of global magma oceans on exoplanet demographics

### 2023-02-23 src/tex/ms.tex
streamline planet categories

    Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
    To enable a quantitative assessment of this detectability, we injected the signal into the simulated planet sample before observing it with simulated surveys.
    While we search for the runaway greenhouse pattern in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's radius based on its individual set of properties and the associated predictions from steam atmosphere and water incorporation models.
    Relevant properties are a planet's mass $M$, its net instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
    We consider three different cases:
    \begin{itemize}
   \item[A.] No radius change.
    This case serves as our null hypothesis, where planetary radii are not affected by the amount of irradiation the planets receive.
    A solid planetary surface is assumed and no steam atmosphere is present.
   \item[B.] Radius inflation due to a steam atmosphere.
    Atmospheric scale heights are significantly increased.
   \item[C.] As B, but including radius decrease of inflated planets due to water partitioning in the molten mantle.
    \end{itemize}

In case~A, a planet simply retains the radius it was assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
There is no additional dependency on orbital distance or instellation.

Case~B assumes an inflated radius due to a steam atmosphere for all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
For the rocky interiors, their calculations rely on the same mass-radius relations from ~\citep{Zeng2016} that we applied for case~A-planets.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
We note that an additional radius change due to potentially molten interiors~\citep{Bower2019} is not taken into account here.

Case~C contains case~B, but also includes a radius decrease due to water partitioning in the melt.
Due to greenhouse forcing, planetary surfaces below steam atmospheres are expected to be molten.
Part of the water will then partition into the melt, removing it from the steam atmosphere, which in turn decreases the radius inflation.
This effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes typically reach measurable values~\citep{Dorn2021}.
The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
We then add the (in almost all cases negative) radius deviations to the planet radii computed for case~B.

\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiuscomparison.pdf}
    \end{centering}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiusevolution.pdf}
        \caption{
            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \wrr$ and different planet states. Green: planets with a solid mantle and no steam atmosphere. Dashed: planets with steam atmospheres. Blue: planets with steam atmospheres and including the effect of water incorporation in the melt.
            \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies among planet classes. Shown is a schematic time evolution of the transit radius normalized to the atmosphere-free radius $R_\mathrm{0}$ for different scenarios. Planets can move between planet classes through processes such as atmospheric loss and desiccation, which ultimately ends a runaway greenhouse phase on a timescale dependent on a planet's water content.}
        \label{fig:radiusevolution}
    \end{centering}
\end{figure}
Figure~\ref{fig:radiusevolution} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \wrr$.
Steam atmospheres (case~B) cause a significant radius increase, which is slightly reduced when water incorporation in the melt is allowed (case~C).
In the following, we only distinguish between cases~A and~C.

As illustrated in Fig.~\ref{fig:radiusevolution}~(bottom panel), different processes can lead to the classification of a planet into either of these categories.
For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a case~C planet, diluting the expected demographic signature.
Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to category~A on a timescale of $10^8\SI{}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
To account for these effects we introduce a dilution parameter $f_\mathrm{rgh}$, which represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.
%The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
This parametrization introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
In the following, we test if and under what conditions this trend is large enough to be detected with high significance.

>>>>

Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
To enable a quantitative assessment of this detectability, we injected the signal into the simulated planet sample before observing it with simulated surveys.
While we search for the runaway greenhouse pattern in demographic quantities such as average planet radii, the injected changes happen on the planetary level: We changed each planet's radius based on its individual set of properties and the associated predictions from steam atmosphere and water incorporation models.
Relevant properties are a planet's mass $M$, its net instellation $S$, and its bulk water inventory expressed as a water mass fraction $x_{H_2O}$.
We consider the following cases (see Fig.~\ref{fig:radiusevolution}):
%\begin{itemize}
%\item[A.] No radius change.
%This case serves as our null hypothesis, where planetary radii are not affected by the amount of irradiation the planets receive.
%A solid planetary surface is assumed and no steam atmosphere is present.
%\item[B.] Radius inflation due to a steam atmosphere.
%Atmospheric scale heights are significantly increased.
%\item[C.] As B, but including radius decrease of inflated planets due to water partitioning in the molten mantle.
%\end{itemize}

\textit{Non-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
This case serves as our null hypothesis.

For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$, we assume an inflated radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
Their calculations rely on the same mass-radius relations from ~\citep{Zeng2016} for rocky interiors that we apply for our non-runaway planets.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
This scenario does not take into account the expected additional radius change due to interiors that are molten due to greenhouse forcing~\citep{Bower2019}.
It further assumes a \textit{dry melt} without dissolved volatiles.

\textit{Wet melt} planets follow the rules above, but we take into account a radius decrease from water partitioning in the melt:
In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
This removes it from the steam atmosphere, which in turn decreases the radius inflation.
The effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes typically reach measurable values~\citep{Dorn2021}.
The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
We then add the (in almost all cases negative) radius deviations to the planet radii computed for the dry melt case.

\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiuscomparison.pdf}
    \end{centering}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/radiusevolution.pdf}
        \caption{
            \textit{Top}: Mass-radius relationships for a water mass fraction $x_{H_2O}= \wrr$ and different planet states. Green: planets with a solid mantle and no steam atmosphere. Dashed: planets with steam atmospheres. Blue: planets with steam atmospheres and including the effect of water incorporation in the melt.
            \textit{Bottom}: Radius evolution of different planet types, illustrating degeneracies and potential for confusion among planet classes. Shown is a schematic time evolution of the transit radius normalized to the atmosphere-free radius $R_\mathrm{0}$ for different scenarios. Planets can move between planet classes through processes such as atmospheric loss and desiccation, which ultimately ends a runaway greenhouse phase on a timescale dependent on a planet's water content.}
        \label{fig:radiusevolution}
    \end{centering}
\end{figure}
Figure~\ref{fig:radiusevolution} shows the different mass-radius relations of the three cases for a fiducial water mass fraction of $x_{H_2O}= \wrr$.
Steam atmospheres cause a significant radius increase, which is slightly reduced when water incorporation in the melt is considered.
In the following, we only distinguish between the non-runaway greenhouse and wet melt scenarios.

As illustrated in Fig.~\ref{fig:radiusevolution}~(bottom panel), different processes can lead to the classification of a planet into either of these categories.
For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the non-runaway greenhouse category on a timescale of $10^8\SI{}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
To account for these effects, we introduce a dilution parameter $f_\mathrm{rgh}$, which represents the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
%The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
This parametrization introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
In the following, we test if and under what conditions this trend is large enough to be detected with high significance.
